---
title: "Severe Weather Consequences in population health and economic aspects"
author: "Rolf Eleveld"
date: "Friday, April 24, 2015"
output: html_document
---
#Severe Weather and Consequences for population health and economy
This research documents the consequences of Severe Weather as collected by NOAA.
Special attention is placed on the population health consequences and the economic impact.
The data is gathered from [NOAA Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) for years 1950-2011.
This research will show what type of events have had greatest impact on population health and which ones had the greatest economic impact. 

#Data Processing
##downloading the data
```{r collecting data, cache=TRUE}
if (!file.exists("NOAA_Severe_Weather_Data.bz2")){
    setInternet2(use = TRUE)
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                  "NOAA_Severe_Weather_Data.bz2",
                  mode="wb")
    }
# unzip the contents of the bzip2 file to a memory table
con <- bzfile('NOAA_Severe_Weather_Data.bz2', 'r')
NOAA <- read.csv(con, as.is=TRUE)
close(con)
```

##processing data
Adding a column where the total value is calculated based on Property and Crop multiplied by the Exp value
```{r combine expense into something that can be compared}
#adding a function, as I could not solve this with a simple grep and sapply...
parseExp <- function(x) {
     exp <- 0
     if (!length(grep("[^[:digit:]]", as.character(x)))) {
         exp <- as.numeric(x); 
     } else {
         if(x=="h" || x=="H"){
             exp <- 2
        } else { if (x=="k" || x=="K") {
            exp<-3
            } else { if (x=="m" || x=="M") {
                exp<-6
                } else { 
                    if (x=="b" || x=="B") {
                        exp<-9
                    }
                }
            }
        } 
     }
     result <- 10^exp
     return(result)
 }
#adding a function to easily process numbers without NA issues
parseNumber <- function(x){
    num <- 0
    if (!is.na(x) && is.numeric(x)){
        num <- as.numeric(x)
    }
    return(num)
}
#adding a column to NOAA with the total damages value
NOAA$TotalDamages <- parseNumber(NOAA$PROPDMG) * parseExp(NOAA$PROPDMGEXP) + 
    parseNumber(NOAA$CROPDMG) * parseExp(NOAA$CROPDMGEXP)
```

##The 10 most costly event types
I take the 10 on average most costly event types and report these here.
```{r 10 most costly event types}
library(plyr)
library(dplyr)
#calculating mean damages
Noaa_damages_by_event <- ddply(NOAA,.(EVTYPE),summarise,Average_Damage=mean(TotalDamages) )
Noaa_top_10_damages <- Noaa_damages_by_event %>%
     arrange(desc(Average_Damage)) %>%
     head(10)
names(Noaa_top_10_damages)<-c("Event_Type", "Average_Damage_per_event")
Noaa_top_10_damages[,1]<-factor(Noaa_top_10_damages[,1])
Noaa_top_10_damages
```

##Comparing injuriesis and deaths is hard
It is hard to sum fatalities and injuries therefore I will consider them separate

###Deadliest Event types
```{r 10 deadliets event types}
Noaa_deaths_by_event <- ddply(NOAA,.(EVTYPE),summarise,Average_Deaths=mean(FATALITIES) )
Noaa_top_10_deaths <- Noaa_deaths_by_event %>%
     arrange(desc(Average_Deaths)) %>%
     head(10)
names(Noaa_top_10_deaths)<-c("Event_Type", "Average_deaths_per_event")
Noaa_top_10_deaths[,1]<-factor(Noaa_top_10_deaths[,1])
Noaa_top_10_deaths
```

###Events with most Injuries
```{r 10 event types with most injuries}
Noaa_injuries_by_event <- ddply(NOAA,.(EVTYPE),summarise,Average_Injuries=mean(INJURIES) )
Noaa_top_10_injuries <- Noaa_injuries_by_event %>%
     arrange(desc(Average_Injuries)) %>%
     head(10)
names(Noaa_top_10_injuries)<-c("Event_Type", "Average_injuries_per_event")
Noaa_top_10_injuries[,1]<-factor(Noaa_top_10_injuries[,1])
Noaa_top_10_injuries
```

#Results
Below is a set of summary charts displaying the impact of the top 10 for:
* Damages
* Deaths
* Injuries

```{r plots}
par(mar=c(8, 4, 2, 2) + 0.1)
par(las=2)
par(cex.axis = 0.5)
barplot(Noaa_top_10_damages$Average_Damage_per_event, 
        names.arg=Noaa_top_10_deaths$Event_Type,
        main="Top 10 average damages", 
        horiz=FALSE)

barplot(Noaa_top_10_deaths$Average_deaths_per_event, 
        names.arg=Noaa_top_10_damages$Event_Type,
        main="Top 10 average deaths", 
        horiz=FALSE)

barplot(Noaa_top_10_injuries$Average_injuries_per_event,
        names.arg=Noaa_top_10_injuries$Event_Type,
        main="Top 10 average injuries", 
        horiz=FALSE)
```

  
No conclusions can be drawn based on these aggregate number over 61 years of archive.
Further study will need to be done to the nature and combination of various weather events.